---
title: "RPR"
author: "Myself"
date: "2024-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

### 1. Load the data (i.e. read.csv())

```{r}
rawdata <- unzip("activity.zip")
activity <- read.csv(rawdata)

head(activity)
```

### 2. Process/transform the data into a format suitable for your data analysis i.e converting string date character into date


``` {r}
activity$date <- as.Date(activity$date,format="%Y-%m-%d")
head(activity)
```
## What is mean total number of steps taken per day?

### 1.Total no.steps taken per day

```{r}
library(dplyr)

total_steps_data <- activity %>%na.omit() %>% group_by(date) %>% summarize(total_steps=sum(steps))
head(total_steps_data)
```
### 2. Histogram for total number steps taken per day
``` {r}
with(total_steps_data,hist(total_steps,breaks=20,xlab='year',ylab='total steps',main="Total no.of steps in a year",col='red'))

```

```{r}
meansteps <- round(mean(total_steps_data$total_steps))
mediansteps <- round(median(total_steps_data$total_steps))

```
The mean and median of the total number of steps in a year `r meansteps` and `r mediansteps`

## What is the average daily activity pattern?

### 1. Calculating the average number of steps/5 minute interval
```{r}
avg_steps_data <- activity %>% na.omit() %>% group_by(interval) %>% summarize(avg_steps=mean(steps))
```
### 2. Plotting a histogram for average steps for 5 minute interval

```{r}
with (avg_steps_data, plot(interval,avg_steps,type="l",xlab="Interval",ylab="Average steps taken",main="Average steps taken during 5 minute interval"))
```

### 3. Finding the 5 minute interval which has max no. of steps

```{r}
intervalWithMaxSteps <- avg_steps_data$interval[which.max(avg_steps_data$avg_steps)]
intervalWithMaxSteps
```

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset

```{r}

missing_values <- colSums(is.na(activity))
missing_values

```
### 2. Fixing the missing values by mean of steps
```{r}
imputed_data <- activity
imputed_data$steps[is.na(imputed_data$steps)] <-  mean(imputed_data$steps,na.rm=T)
missing_values <- colSums(is.na(imputed_data))
missing_values
```
```{r}
imputed_data$date <- as.Date(imputed_data$date,format="%Y-%m-%d")
head(imputed_data)
```

```{r}
total_steps_imputed <- imputed_data %>% group_by(date) %>% summarize(total_steps_imp=sum(steps))
with(total_steps_imputed,hist(total_steps_imp,breaks=20,xlab="Total number of steps taken each day",ylab="Count",main="Total number of steps taken each day",col="blue"))
```
```{r}
imputed_meansteps <- round(mean(total_steps_imputed$total_steps_imp))
imputed_mediansteps <- round(median(total_steps_imputed$total_steps_imp))
imputed_meansteps
imputed_mediansteps
```
### Inference:
#### The difference between means for raw data and imputed data is `r imputed_meansteps - meansteps `.

#### The difference between medians for raw data and imputed data is `r imputed_mediansteps - mediansteps`.

#### By imputing the 'NA' values with mean of steps in the raw data ,  the mean remained the same but the median has changed marginally by 1 and moved closer to the mean of the raw data.

# Are there differences in activity patterns between weekdays and weekends?

## 1.Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
imputed_data$day <- weekdays(imputed_data$date)

imputed_data$dayType <- "weekday"

imputed_data$dayType[imputed_data$day %in% c("Saturday","Sunday")] <- "weekend"


```

```{r}
weekData <- imputed_data %>% group_by(dayType,interval) %>% summarize(avg_day_steps=mean(steps))




```
## 2. Make a panel plot containing a time series plot for 5-minute interval (x-axis) and the average number of steps taken

```{r}

library(ggplot2)

qplot(interval,avg_day_steps,data=weekData, geom="line",xlab="Interval",ylab="Number of steps (Average)", facets=dayType ~ .)

```

### Inference : The average steps is more in the weekdays as per the plot which may be related to people working in weekdays.
